\textbf{The Structure of Chaos: An Empirical Comparison of Fractal Physiology Complexity Indices using NeuroKit2}

### 2022-07-20 manuscript.tex
R1 - complete

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{chaos, complexity, fractal, physiology, noise\newline\indent Word count: 3079}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

\abstract{%
Complexity quantification, through entropy, information theory and fractal dimension indices, is gaining a renewed traction in psychophsyiology, as new measures with promising qualities emerge from the computational and mathematical advances. Unfortunately, few studies compare the relationship and objective performance of the plethora of existing metrics, in turn hindering reproducibility, replicability, consistency, and clarity in the field. Using the NeuroKit2 Python software, we computed a list of 112 complexity indices on signals varying in their characteristics (noise, length and frequency spectrum). We then systematically compared the indices by their computational weight, their representativeness of a multidimensional space of latent dimensions, and empirical proximity with other indices. Based on these considerations, we propose that a selection of 12 indices, together representing 85.97\% of the total variance of all indices, might offer a parsimonious and complimentary choice in regards to the quantification of the complexity of time series. Our selection includes \emph{CWPEn}, \emph{Line Length (LL)}, \emph{BubbEn}, \emph{MSWPEn}, \emph{MFDFA (Max)}, \emph{Hjorth Complexity}, \emph{SVDEn}, \emph{MFDFA (Width)}, \emph{MFDFA (Mean)}, \emph{MFDFA (Peak)}, \emph{MFDFA (Fluctuation)}, \emph{AttEn}. Elements of consideration for alternative subsets are discussed, and data, analysis scripts and code for the figures are open-source.
}

@@

\begin{figure}
\centering
\includegraphics{./figures/time-1.pdf}
\caption{\label{fig:time}Median computation time difference between the different complexity indices algorithms, as well as variability as a function of signal lengths (represented by different line colors). The indices are grouped in sections (background color) according to their median computation time. Note that the time is expressed in arbitrary units as it is intended to convey differences, as the actual time would depend on the system specifications.}
\end{figure}

@@

The second factor was strongly loaded by by indices that measure the feature-richeness of the signal's system (as most of them operate on a state-space decomposition). It was found to be positively related to \emph{SVDEn} and the Kozachenko-Leonenko differential entropy (\emph{KLEn}), and negatively to the RQA \emph{Recurrence Rate} and \emph{Hjorth} Complexity.

The third factor was loaded predominantly by permutation-based metrics (\emph{PEn}, \emph{WPEn}, \emph{BubblEn}, etc.). The fourth factor included multiscale indices, such as \emph{MSWPEn}. The fifth factor was strongly loaded by signal \emph{length}, and thus might not capture features of complexity \emph{per se}. Indices with the most relation to it were indices generally known to be sensitive to signal length, such as \emph{ApEn}. The sixth factor was loaded by indices in which the signal or the Poincaré plot was discretized via binning or gridding, respectively. The seventh factor was loaded by sign-based entropy increments, and the eights by multiscale \emph{IncrEn} and multiscale \emph{PLZC}. The ninth factor was loaded by \emph{EnofEn} and Kolmogorov Entropy (\emph{K2En}). The tenth factor was loaded positively by the amount of noise, and negatively by multifractal indices such as \emph{MFDFA (Width)}, suggesting a sensitivity to regularity. Finally, as a manipulation check for our factorization method, the random vector did not load unto any factors.

@@

For illustration purposes, we represented the correlation matrix as a connectivity graph (see \textbf{Figure 5}). We then ran a hierarchical clustering (with a Ward D2 distance) to provide additional information about the groups discussed above. Indeed, while the factor analysis will predominantly show indices that are the most representative of a given latent dimension, clustering will construct groups based on the multidimensional profile (what dimensions a given index loads positively on, and what other does it load negatively on). This allowed us to refined our recommendations for complimentary complexity indices (see \textbf{Figure 6}).

@@

The selection of a subset of indices was based on a set of considerations: 1) high loadings on one predominant latent dimension, with additional attention to the pattern of secondary loadings. For instance, an index with a positive factor 1 loading and a negative factor 2 loading could complement another index with a similar factor 1 loading, but a positive factor 2 loading. This was facilitated by 2) the hierarchical clustering dendrogram (see \textbf{Figure 6}), with which we attempted to extract indices from each (meaningful) higher order clusters. Items related to clusters that we determine as being largely explained by noise, length or other artifacts were omitted. 3) A preference for indices with relatively shorter computation times. This yielded a selection of 12 indices. Next, we computed the cumulative variance explained of this selection in respect to the entirety of indices examined, and derived the optimal order to maximize the variance explained (see \textbf{Figure 7}). The 12 included indices, representing 85.97\% of the variance of the whole dataset, were:

\begin{itemize}
\tightlist
\item
  \emph{CWPEn}: The Conditional Weighted Permutation Entropy is based on the difference of weighted entropy between that obtained at an embedding dimension \(m\) and that obtained at \(m+1\) (Unakafov \& Keller, 2014).
\item
  \emph{LL}: The Line Length index stems out of a simplification of Katz' fractal dimension (\emph{KFD}) algorithm (Esteller et al., 2001) and corresponds to the average of consecutive absolute differences. It is equivalent to \emph{NDLFD}, the Fractal dimension via Normalized Length Density (Kalauzi et al., 2009). As it captures the amplitude 1-lag fluctuations, this index is likely sensitive to noise in the series.
\item
  \emph{BubbEn}: The Bubble Entropy is based on Permutation Entropy. It uses the \emph{Bubble sort} algorithm and counts the number of swaps each vector undergoes in the embedding space instead of ranking their order ({``Bubble Entropy,''} 2017).
\item
  \emph{MSWPEn}: The Multiscale Weighted Permutation Entropy is the entropy of weighted ordinal descriptors of the time-embedded signal computed at different scales obtained by a coarse-graining procedure (Fadlallah et al., 2013).
\item
  \emph{MFDFA (Max)} : The value of singularity spectrum \emph{D} corresponding to the maximum value of singularity exponent \emph{H}.
\item
  \emph{Hjorth}: Hjorth's Complexity is defined as the ratio of the mean frequency of the first derivative of the signal to the mean frequency of the signal (Hjorth, 1970).
\item
  \emph{SVDEn}: The Singular Value Decomposition (SVD) Entropy quantifies the amount of eigenvectors needed for an adequate representation of the system (Roberts et al., 1999).
\item
  \emph{MFDFA (Width)}: The width of the multifractal singularity spectrum (Kantelhardt et al., 2002) obtained via Detrended Fluctuation Analysis (DFA).
\item
  \emph{MFDFA (Mean)} : The mean of the maximum and minimum values of singularity exponent \emph{H}, which quantifies the average fluctuations of the signal.
\item
  \emph{MFDFA (Peak)} : The value of the singularity exponent \emph{H} corresponding to peak of singularity dimension \emph{D}. It is a measure of the self-affinity of the signal, and a high value is an indicator of high degree of correlation between the data points.
\item
  \emph{MFDFA (Increment)}: The cumulative function of the squared increments of the generalized Hurst's exponents between consecutive moment orders (Faini et al., 2021).
\item
  \emph{AttEn}: The Attention Entropy is based on the frequency distribution of the intervals between the local maxima and minima of the time series (J. Yang et al., 2020).
\end{itemize}

@@

As the span and application of complexity science grows, a systematic approach to compare their ``performance'' becomes necessary to reinforce the clarity and structure of the field. The term \emph{performance} used here is to be understood in a relative sense, as any such endeavor faces the ``hard problem'' of complexity science: various objective properties of signals (e.g., short-term vs.~long-term variability, auto-correlation, information, randomness, Namdari \& Li, 2019; Xiong et al., 2017) participate in forming overarching concepts such as ``complex'' and ``chaotic''. Indices that are sensitive to some of these objective properties are thus conceptually linked through such over-arching frameworks. However, it remains unclear how these high-level concepts transfer back, in a top-down fashion, into a combination of lower-level features. As such, it is conceptually complicated to benchmark complexity measures against ``objectively'' complex vs.~non-complex signals. In other words, while we are aware that different objective signal characteristics can contribute to the ``complexity'' of a signal, there is not a one-to-one correspondence between the latter and the former.

@@

\leavevmode\vadjust pre{\hypertarget{ref-bubblee2017}{}}%
Bubble entropy: An entropy almost free of parameters. (2017). \emph{IEEE Transactions on Biomedical Engineering}, \emph{64}(11), 2711--2718. \url{https://doi.org/10.1109/tbme.2017.2664105}

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{chaos, complexity, fractal, physiology, noise\newline\indent Word count: 3140}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

\abstract{%
Complexity quantification, through entropy, information theory and fractal dimension indices, is gaining a renewed traction in psychophsyiology, as new measures with promising qualities emerge from the computational and mathematical advances. Unfortunately, few studies compare the relationship and objective performance of the plethora of existing metrics, in turn hindering reproducibility, replicability, consistency, and clarity in the field. Using the NeuroKit2 Python software, we computed a list of 112 (predominantly used) complexity indices on signals varying in their characteristics (noise, length and frequency spectrum). We then systematically compared the indices by their computational weight, their representativeness of a multidimensional space of latent dimensions, and empirical proximity with other indices. Based on these considerations, we propose that a selection of 12 indices, together representing 85.97\% of the total variance of all indices, might offer a parsimonious and complimentary choice in regards to the quantification of the complexity of time series. Our selection includes \emph{CWPEn}, \emph{Line Length (LL)}, \emph{BubbEn}, \emph{MSWPEn}, \emph{MFDFA (Max)}, \emph{Hjorth Complexity}, \emph{SVDEn}, \emph{MFDFA (Width)}, \emph{MFDFA (Mean)}, \emph{MFDFA (Peak)}, \emph{MFDFA (Fluctuation)}, \emph{AttEn}. Elements of consideration for alternative subsets are discussed, and data, analysis scripts and code for the figures are open-source.
}

@@

\begin{figure}
\centering
\includegraphics{./figures/time-1.pdf}
\caption{\label{fig:time}Median computation time difference between the different complexity indices algorithms, as well as variability as a function of signal lengths (represented by different line colors). The indices are grouped in sections (background color) according to their median computation time. Note that the time is expressed in arbitrary units as it is intended to convey differences, since the actual time would depend on the system specifications.}
\end{figure}

@@

The second factor was strongly loaded by indices that measure the feature-richness of the signal's system (as most of them operate on a state-space decomposition). It was found to be positively related to \emph{SVDEn} and the Kozachenko-Leonenko differential entropy (\emph{KLEn}), and negatively to the RQA \emph{Recurrence Rate} and \emph{Hjorth} Complexity.

The third factor was loaded predominantly by permutation-based metrics (\emph{PEn}, \emph{WPEn}, \emph{BubblEn}, etc.). The fourth factor included multiscale indices, such as \emph{MSWPEn}. The fifth factor was strongly loaded by signal \emph{length}, and thus might not capture features of complexity \emph{per se}. Indices with the most relation to it were indices generally known to be sensitive to signal length, such as \emph{ApEn}. The sixth factor was loaded by indices in which the signal or the Poincaré plot was discretized via binning or gridding, respectively. The seventh factor was loaded by sign-based entropy increments, and the eighth by multiscale \emph{IncrEn} and multiscale \emph{PLZC}. The ninth factor was loaded by \emph{EnofEn} and Kolmogorov Entropy (\emph{K2En}). The tenth factor was loaded positively by the amount of noise, and negatively by multifractal indices such as \emph{MFDFA (Width)}, suggesting a sensitivity to regularity. Finally, as a manipulation check for our factorization method, the random vector did not load unto any factors.

@@

For illustration purposes, we represented the correlation matrix as a connectivity graph (see \textbf{Figure 5}). We then ran a hierarchical clustering (with a Ward D2 distance) to provide additional information about the groups discussed above. Indeed, while the factor analysis will predominantly show indices that are the most representative of a given latent dimension, clustering will construct groups based on the multidimensional profile (what dimensions a given index loads positively on, and what other does it load negatively on). This allowed us to refine our recommendations for complimentary complexity indices (see \textbf{Figure 6}).

@@

The selection of a subset of indices was based on a set of considerations: 1) high loadings on one predominant latent dimension, with additional attention to the pattern of secondary loadings. For instance, an index with a positive factor 1 loading and a negative factor 2 loading could complement another index with a similar factor 1 loading, but a positive factor 2 loading. This was facilitated by 2) the hierarchical clustering dendrogram (see \textbf{Figure 6}), with which we attempted to extract indices from each (meaningful) higher order clusters. Items related to clusters that we determine as being largely explained by noise, length or other artifacts were omitted. 3) A preference for indices with relatively shorter computation times. This yielded a selection of 12 indices. Next, we computed the cumulative variance explained by this selection in respect to the entirety of indices examined, and derived the optimal order to maximize the variance explained (see \textbf{Figure 7}). The 12 included indices, representing 85.97\% of the variance of the whole dataset, were:

\begin{itemize}
\tightlist
\item
  \emph{CWPEn}: The Conditional Weighted Permutation Entropy is based on the difference of weighted entropy between that obtained at an embedding dimension \(m\) and that obtained at \(m+1\) (Unakafov \& Keller, 2014).
\item
  \emph{LL}: The Line Length index stems out of a simplification of Katz' fractal dimension (\emph{KFD}) algorithm (Esteller et al., 2001) and corresponds to the average of consecutive absolute differences. It is equivalent to \emph{NDLFD}, the Fractal dimension via Normalized Length Density (Kalauzi et al., 2009). As it captures the amplitude 1-lag fluctuations, this index is likely sensitive to noise in the series.
\item
  \emph{BubbEn}: The Bubble Entropy is based on Permutation Entropy. It uses the \emph{Bubble sort} algorithm and counts the number of swaps each vector undergoes in the embedding space instead of ranking their order (Manis et al., 2017).
\item
  \emph{MSWPEn}: The Multiscale Weighted Permutation Entropy is the entropy of weighted ordinal descriptors of the time-embedded signal computed at different scales obtained by a coarse-graining procedure (Fadlallah et al., 2013).
\item
  \emph{MFDFA (Max)} : The value of singularity spectrum \emph{D} corresponding to the maximum value of singularity exponent \emph{H}.
\item
  \emph{Hjorth}: Hjorth's Complexity is defined as the ratio of the mean frequency of the first derivative of the signal to the mean frequency of the signal (Hjorth, 1970).
\item
  \emph{SVDEn}: The Singular Value Decomposition (SVD) Entropy quantifies the amount of eigenvectors needed for an adequate representation of the system (Roberts et al., 1999).
\item
  \emph{MFDFA (Width)}: The width of the multifractal singularity spectrum (Kantelhardt et al., 2002) obtained via Detrended Fluctuation Analysis (DFA).
\item
  \emph{MFDFA (Mean)} : The mean of the maximum and minimum values of singularity exponent \emph{H}, which quantifies the average fluctuations of the signal.
\item
  \emph{MFDFA (Peak)} : The value of the singularity exponent \emph{H} corresponding to peak of singularity dimension \emph{D}. It is a measure of the self-affinity of the signal, and a high value is an indicator of high degree of correlation between the data points.
\item
  \emph{MFDFA (Increment)}: The cumulative function of the squared increments of the generalized Hurst's exponents between consecutive moment orders (Faini et al., 2021).
\item
  \emph{AttEn}: The Attention Entropy is based on the frequency distribution of the intervals between the local maxima and minima of the time series (J. Yang et al., 2020).
\end{itemize}

@@

As the span and application of complexity science grows, a systematic approach to compare their ``performance'' becomes necessary to reinforce the clarity and structure of the field. The term \emph{performance} used here is to be understood in a relative sense, as any such endeavor faces the ``hard problem'' of complexity science: various objective properties of signals (e.g., short-term vs.~long-term variability, auto-correlation, information, randomness, Namdari \& Li, 2019; Xiong et al., 2017) participate in forming overarching concepts such as ``complex'' and ``chaotic''. Indices that are sensitive to some of these objective properties are thus conceptually linked through such overarching frameworks. However, it remains unclear how these high-level concepts transfer back, in a top-down fashion, into a combination of lower-level features. As such, it is conceptually complicated to benchmark complexity measures against ``objectively'' complex vs.~non-complex signals. In other words, while we are aware that different objective signal characteristics can contribute to the ``complexity'' of a signal, there is not a one-to-one correspondence between the latter and the former.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-bubblee2017}{}}%
Manis, G., Aktaruzzaman, M., \& Sassi, R. (2017). Bubble entropy: An entropy almost free of parameters. \emph{IEEE Transactions on Biomedical Engineering}, \emph{64}(11), 2711--2718. \url{https://doi.org/10.1109/tbme.2017.2664105}
